---
title: ""
author: "Mason Shihab"
date: "3/13/2021"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lubridate)
library(tidyverse)
library(ggplot2)
library(readr)
library(tidycensus)
library(rpart)
library(rpart.plot)
library(ipred)
library(readxl)
library(usdata)
library(tm)
library(stringr)
```


# Introduction

Today we will be expanding on the election data set and partition it between a test and a train data set. We will use the train data set in order to attempt to predict the remainder of the values. Various methods for doing so will be ran and compared against each other for educational and practice purposes.



## Reloading the data.
```{r}
var <- load_variables(2019, "acs1", cache = TRUE)
```


## Selecting the variables

```{r}
census_data <- get_acs(geography = "county", variables = c(med_age = "B01002_001", 
                                                           permale = "B01001_002",
                                                           bachplus = "B16010_041",
                                                           totalpop = "B01003_001",
                                                           unemployed = "B27011_008",
                                                           employed =	"B27011_003",
                                                           foodstamp = "B09010_002",
                                                           ilefnhi= "B27011_007",
                                                           ilufnhi= "B27011_012",
                                                           nilfnhi = "B27011_017",
                                                           gini = "B19083_001",
                                                           med_income = "B19013_001",
                                                           med_2bed = "B25031_004",
                                                           single_mom = "B11012_010",
                                                           lessthan_hs = "B16010_002",
                                                           housing_units = "B25001_001"), 
                       year = 2019)





census_step = census_data %>% 
  pivot_wider(id_cols = c(NAME), names_from = variable, values_from = estimate) %>% 
  separate(NAME, into = c("county", "state"), sep = ", ") %>% arrange(state, county)


census_wider = census_step %>% 
  mutate(permale = permale/totalpop, 
         bachplus = bachplus/totalpop, unemployed_rate = unemployed/(employed+unemployed), employed_rate = employed/(employed+unemployed),
         foodstamp = foodstamp/totalpop, no_health_ins = 
           (ilefnhi+ilufnhi+nilfnhi)/totalpop, single_mom = 
           single_mom/totalpop, lessthan_hs = lessthan_hs/totalpop) %>% 
  select(-ilefnhi,-ilufnhi,-nilfnhi, -unemployed, -employed)
ACS_vars = drop_na(census_wider)



```



## Combine with Election Dataset

```{r}
election = read_csv("president_county_candidate.csv", 
    col_types = cols(won = col_skip()))



election2 = election %>% arrange(state, county) %>% 
  group_by(state, county) %>% 
  mutate(pctvote = 100*total_votes/sum(total_votes)) %>% 
  filter(candidate %in% c("Joe Biden", "Donald Trump"))

election_clean = election2 %>% 
  pivot_wider(id_cols = c(state, county), 
              names_from = candidate, values_from = pctvote)

ACS_election = election_clean %>% inner_join(ACS_vars, by = c("state", "county")) %>% select(-`Joe Biden`) %>% rename(pertrump = 'Donald Trump') %>% mutate(pertrump = pertrump/100)

```

```{r}
keyfile = read_csv("../data/raw/ZIP-COUNTY-FIPS_2010-03.csv") %>% 
  mutate(state_full = abbr2state(STATE),) %>% rename(county = COUNTYNAME, state = state_full, fips = STCOUNTYFP) %>% select(-ZIP) %>% distinct() 
```

```{r}
ACS_election_key = ACS_election %>% inner_join(keyfile, by = c("state", "county")) 
```


## Land and Density

```{r}
land = read_xlsx("land.xlsx") %>% rename(fips = STCOU, area = LND010190D) %>% mutate(fips = as.numeric(fips)) 
```


```{r}
ACS_election_land = ACS_election_key %>% inner_join(land, by = "fips")
```

```{r}
ACS_election_land = ACS_election_land %>% mutate(pop_density = totalpop/area, housing_density = housing_units/area)
```