---
title: "EDA"
author: "Elinor chu"
date: "12/6/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lubridate)
library(tidyverse)
library(ggplot2)
library(readr)
library(readxl)
library(usdata)
library(tm)
library(stringr)
library(kableExtra)
library(corrplot)
```

```{r}
theft_alldata=read_csv("../data/clean/dataclean.csv")
```


```{r}
theft_train=read_csv("../data/clean/theft_train.csv")
```



# EDA - Response Variable

## Histogram of Response

```{r}
theft_train %>% ggplot(aes(x = theftrate)) +
  geom_histogram()+
  labs(y = "Count", 
       x = "Theft Rate",
       title = "Histogram of Theft Rate")+
  geom_vline(aes(xintercept = mean(theftrate),colour = "mean"),
             linetype ="longdash", size = .8)+
  geom_vline(aes(xintercept = median(theftrate),colour = "median"),
             linetype ="longdash", size = .8)+
  theme_bw()+
  scale_y_log10()+
  scale_color_manual(name = "statistics", values = c(median = "blue", mean = "red"))
```
```{r}
mean(theft_train$theftrate)
median(theft_train$theftrate)
```


## Boxplot of Response

```{r}
theft_train %>% select(-fips, -state, -county) %>%
ggplot(aes(y=theftrate)) +
  geom_boxplot()+
  labs(y = "Theft Rate",
       title = "Boxplot of Theft Rate")+
  theme_bw()
```

To understand the distribution of the response variable, we first ploted a histogram of theft rate. As seen from Figure X, the data appears to be right-skewed, with some counties exceeding a theft rate of 4%. The mean county-level theft rate in 2020 is 0.563%; the median is 0.417%. The boxplot in Figure X provides a clearer presentation of the outlier counties with exceedingly high theft rates. 

## Highest Theft Rate - Top10 Counties

```{r}
theft_train %>% select(state,county,theftrate) %>% arrange(desc(theftrate)) %>% head(10) %>%
  kable(format = "latex", row.names = NA, 
        booktabs = TRUE,
        digits = 3,
        col.names = c("State", "County","Theft Rate"),
        caption = "This a table showing the top 10 counties with the highest theft rate.") %>%
  kable_styling(position = "center") %>%
  kable_styling(latex_options = "HOLD_position")
```
We proceeded to determine which counties had extremely high theft rate in 2020 by looking at the sorted data. The sorted data in Table X shows that the top 6 counties with highest rates of theft incidence are Hamilton county in IN, New York county in NY, Jefferson county in CO, Denver county in CO, Tunica county in MS, and San Francisco county in CA. 


## Heat map of theft rate

```{r}
map_data("county") %>%
  as_tibble() %>% 
  left_join(theft_train %>% 
              rename(region = state, 
                     subregion = county,
                     `Theft Rate` = theftrate) %>% 
              mutate(region = str_to_lower(region), 
                     subregion = str_to_lower(subregion)), 
            by = c("region", "subregion")) %>%
  ggplot() + 
  geom_polygon(data=map_data("state"), 
               aes(x=long, y=lat, group=group),
               color="black", fill=NA,  size = 1, alpha = .3) + 
  geom_polygon(aes(x=long, y=lat, group=group, fill = `Theft Rate`),
               color="darkblue", size = .1) +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_void()
```

We also created a heat map for our response variable. As shown in Figure X, most of the counties included in this dataset have theft rates well below 2.5%. Very few states have theft rates about 7.5%. The grey areas indicate counties not included in our dataset.



#EDA for important features

## corrplots of all features
```{r}
theft_train_corrAll = theft_train%>% select(-fips, -state, -county,-theftrate)
M = cor(theft_train_corrAll)
corrplot(M, type = 'lower', order = 'hclust', tl.col = 'black',
         cl.ratio = 0.2, tl.srt = 45, col = COL2('PuOr', 10), tl.cex = 0.4)
```

## corrplots of 5 clusters of features

```{r}
cluster_safetynet = theft_train%>% select(-fips, -state, -county) %>% select(unemp_bens_possible, spend_per_capita, saversperhouses, no_health_ins, foodstamp)

cluster_criminaljustice = theft_train%>% select(-fips, -state, -county) %>% select(incar_rate, police_violence_score, police_accountability_score, approach_to_policing_score, police_funding_score)

cluster_health = theft_train%>% select(-fips, -state, -county) %>% select(mean_hha_score, poor_fair_health, dis5to17, dis18to34, dis35to64)

cluster_ses = theft_train%>% select(-fips, -state, -county) %>% select(lessthan_hs, bachplus, unemployed_rate, employed_rate, med_2bed, gini, svi_overall, pct_all_in_pov, pct_child_in_pov, med_income, sev_hou_cost_burden, sev_hou_prob, PctEmpChange1920, PctEmpConstruction, PctEmpMining, PctEmpTrade, PctEmpTrans, PctEmpInformation, PctEmpFIRE, PerCapitaInc, Deep_Pov_All, Deep_Pov_Children, inschool, ingradprofesh, inundergrad)

cluster_demo= theft_train%>% select(-fips, -state, -county) %>% select(med_age,permale,divorced,widowed,nevermarried,pertrump,pop_density,housing_density,res_seg_nonwhite_white,Marriedcouplefamily,ForeignBornEuropePct,ForeignBornMexPct,ForeignBornCaribPct,ForeignBornCentralSouthAmPct,ForeignBornAsiaPct,ForeignBornAfricaPct,NonEnglishHHNum,AvgHHSize,PopChangeRate1819,withkids,single_mom,singledad,foreignborn,fromdifstate,fromabroad)
```

```{r}
M_safetynet = cor(cluster_safetynet)
M_criminaljustice = cor(cluster_criminaljustice)
M_health = cor(cluster_health)
M_ses = cor(cluster_ses)
M_demo = cor(cluster_demo)

```

## social safety net
```{r}
corrplot(M_safetynet, type = 'lower', order = 'hclust', tl.col = 'black',
         cl.ratio = 0.2, tl.srt = 45, col = COL2('PuOr', 10), tl.cex = 1)
```
We observed a positively correlation between State and local government spending on people and State unemployment insurance. Also, Percent of households qualifying for food stamps is positively correlated with No Health Insurance. 
Not surprisingly, No Health Insurance is negatively correlated with State and local government spending on people.


## criminal justice response

```{r}
corrplot(M_criminaljustice, type = 'lower', order = 'hclust', tl.col = 'black',
         cl.ratio = 0.2, tl.srt = 45, col = COL2('PuOr', 10), tl.cex = 0.8)
```
We found no significant correlations among features belonging to the category of Criminal justice response.

##health-related factors

```{r}
corrplot(M_health, type = 'lower', order = 'hclust', tl.col = 'black',
         cl.ratio = 0.2, tl.srt = 45, col = COL2('PuOr', 10), tl.cex = 0.8)
```
We observed positive correlations between Percent adults reporting poor or fair health and Percent of people with disability in each of the three age groups. Moreover, the percentages of people with disability for the three age groups are also positively correlated with each other.

## SES

```{r}
corrplot(M_ses, type = 'lower', order = 'hclust', tl.col = 'black',
         cl.ratio = 0.2, tl.srt = 45, col = COL2('PuOr', 10), tl.cex = 0.6)
```

We observed that there are positive correlations between the Social Vulnerability Index (SVI) and all poverty-related features. We also found that a group of features, including Employment rate, Percent with college or higher education, Median household income, and Per capita income in the past 12 months, that are negatively correlated with SVI and poverty-related features.


## basic demographics

```{r}
corrplot(M_demo, type = 'lower', order = 'hclust', tl.col = 'black',
         cl.ratio = 0.2, tl.srt = 45, col = COL2('PuOr', 10), tl.cex = 0.5)
```

From Figure X, we observed that Percent of household with own children is negatively correlated with 

## Histogram for the Top7 important features (overlaps of Ridge & Lasso)

```{r}
# plot theftrate against poor_fair_health
p1 = theft_train %>% select(-fips, -state, -county) %>% 
  ggplot(aes(x = poor_fair_health, y = theftrate)) +
  geom_point(alpha = 0.6) + 
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", formula = "y~x", se = FALSE) +
  labs(x = "Percent adults reporting poor or fair health", 
       y = "Theft Rate",
       title = "Percentage of Adults Reporting\n Fair or Poor Health vs Theft Rate") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5))

p1
```

```{r}
theft_train %>% arrange(desc(poor_fair_health))%>%head(5)
```



```{r}
# plot theftrate against housing_density
p2 = theft_train %>% select(-fips, -state, -county)%>%
  ggplot(aes(x = housing_density, y = theftrate)) +
  geom_point(alpha = 0.6) + 
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", formula = "y~x", se = FALSE) +
  labs(x = "Housing Density", 
       y = "Theft Rate",
       title = "Housing Density vs Theft Rate") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5))

p2
```


```{r}
theft_train %>% arrange(desc(housing_density))%>%head(5)
```


```{r}
# plot theftrate against pop_density
p3 = theft_train %>% select(-fips, -state, -county) %>%
  ggplot(aes(x = pop_density, y = theftrate)) +
  geom_point(alpha = 0.6) + 
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", formula = "y~x", se = FALSE) +
  labs(x = "Population Density", 
       y = "Theft Rate",
       title = "Population Density vs Theft Rate") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5))

p3
```

```{r}
theft_train %>% arrange(desc(pop_density))%>%head(5)
```


```{r}
# plot theftrate against pertrump
p4 = theft_train %>% select(-fips, -state, -county)%>%
  ggplot(aes(x = pertrump, y = theftrate)) +
  geom_point(alpha = 0.6) + 
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", formula = "y~x", se = FALSE) +
  labs(x = "Percent of Trump Supporters", 
       y = "Theft Rate",
       title = "Percent of Trump Supporters (in 2020) vs Theft Rate") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5))

p4
```

```{r}
theft_train %>% arrange(desc(pertrump))%>%head(5)
```


```{r}
# plot theftrate against PctEmpFIRE
p5 = theft_train %>% select(-fips, -state, -county)%>%
  ggplot(aes(x = PctEmpFIRE, y = theftrate)) +
  geom_point(alpha = 0.6) + 
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", formula = "y~x", se = FALSE) +
  labs(x = "Percent of People Employed in FIRE", 
       y = "Theft Rate",
       title = "Percent of People Employed in \n Finance/Insurance/Real Estate(FIRE) vs Theft Rate") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5))

p5
```

```{r}
theft_train %>% arrange(desc(PctEmpFIRE))%>%select(state, county, PctEmpFIRE, theftrate)%>%head(5)
```


```{r}
# plot theftrate against saversperhouses
p6 = theft_train %>% select(-fips, -state, -county)%>%
  ggplot(aes(x = saversperhouses, y = theftrate)) +
  geom_point(alpha = 0.6) + 
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", formula = "y~x", se = FALSE) +
  labs(x = "Percent of people qualifying for Savers Credit", 
       y = "Theft Rate",
       title = "Percent of people qualifying for\n Saver's Credit vs Theft Rate") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5))

p6
```


```{r}
theft_train %>% arrange(desc(saversperhouses))%>%select(state, county, saversperhouses, theftrate)%>%head(5)
```


```{r}
# plot theftrate against unemp_bens_possible
p7 = theft_train %>% select(-fips, -state, -county)%>%
  ggplot(aes(x = unemp_bens_possible, y = theftrate)) +
  geom_point(alpha = 0.6) + 
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", formula = "y~x", se = FALSE) +
  labs(x = "State unemployment insurance", 
       y = "Theft Rate",
       title = "State Unemployment Insurance vs Theft Rate") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5))

p7
```



```{r}
# plot theftrate against lessthan_hs
p8= theft_train %>% select(-fips, -state, -county)%>%
  ggplot(aes(x = lessthan_hs, y = theftrate)) +
  geom_point(alpha = 0.6) + 
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", formula = "y~x", se = FALSE) +
  labs(x = "Percent with less than HS education", 
       y = "Theft Rate",
       title = "Percent of people with Less than High School\n Education vs Theft Rate") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5))

p8
```

```{r}
theft_train %>% arrange(desc(lessthan_hs))%>%select(state, county, lessthan_hs, theftrate)%>%head(5)
```


```{r}
# plot theftrate against PctEmpConstruction
p9= theft_train %>% select(-fips, -state, -county)%>%
  ggplot(aes(x = PctEmpConstruction, y = theftrate)) +
  geom_point(alpha = 0.6) + 
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", formula = "y~x", se = FALSE) +
  labs(x = "Percent employed in construction", 
       y = "Theft Rate",
       title = "Percent of People Employed in Construction vs. Theft Rate") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5))

p9
```


```{r}
theft_train %>% arrange(desc(PctEmpConstruction))%>%select(state, county, PctEmpConstruction, theftrate)%>%head(5)
```


