---
title: ""
author: "Mason Shihab"
date: "3/13/2021"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lubridate)
library(tidyverse)
library(ggplot2)
library(readr)
library(readxl)
library(usdata)
library(tm)
library(stringr)
library(gbm)
library(glmnetUtils) # boosting
```

```{r}
# install.packages("scales")              # dependency of plot_glmnet
source("functions/plot_glmnet.R")
```

```{r}
theft_train = read.csv('../data/clean/theft_train.csv')
```


# Elastic net regression

Next, let's run an elastic net regression. We can do this via the `cva.glmnet()` function:
```{r}
elnet_fit = cva.glmnet(theftrate ~ . -fips -county -state, # formula notation, as usual
                       nfolds = 10,               # number of folds
                       data = theft_train)   # data to run on
```

The following are the values of `alpha` that were used:
```{r}
elnet_fit$alpha
```

We can plot the minimum CV error for each value of alpha using the helper function `plot_cva_glmnet()` from `plot_glmnet.R`:
```{r}
plot_cva_glmnet(elnet_fit)
```

We can then extract the `cv.glmnet` fit object based on the optimal `alpha` using `extract_best_elnet` from `plot_glmnet.R`:
```{r}
elnet_fit_best = extract_best_elnet(elnet_fit)
```

The `elnet_fit_best` object is a usual `glmnet` fit object, with an additional field called `alpha` specifying which value of `alpha` was used:
```{r}
elnet_fit_best$alpha
```


We can make a CV plot to select `lambda` as usual:
```{r}
plot(elnet_fit_best)
```

And we can make a trace plot for this optimal value of `alpha`:
```{r}
plot_glmnet(elnet_fit_best, theft_train)
```

This is too many features to highlight, so let's choose a smaller number:
```{r}
plot_glmnet(elnet_fit_best, theft_train, features_to_plot = 10)
```


```{r lasso-nonzero-std-coefficients-table}
extract_std_coefs(elnet_fit_best, theft_train) %>% 
  filter(coefficient != 0) %>% arrange(desc(abs(coefficient)))
```
